 

### 2023-02-25 conclusion.Rmd
Abstract and Introduction for Paper 3

---
title: ''
output: pdf_document

---

\addcontentsline{toc}{section}{Conclusion}
\section*{Conclusion}

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

\cite{ngTracingPathChikungunya2010}

>>>>

\section*{Conclusion}

\addcontentsline{toc}{section}{Conclusion}

Overview of Paper 1

Overview of Paper 2

Overview of Paper 3

Policy implications and future research

Burden may be larger than traditionally conceptualized

Papers also show that mental health is linked to broader contextual challenges (climate change, COVID-19) – resilience/vulnerability, exacerbating existing disparities, but also could be critical to overcoming them…

In the absence of population mental health metrics, innovative uses of health data may be helpful in conducting research, but come with limitations

### 2023-02-25 intro.Rmd
Abstract and Introduction for Paper 3

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Lacinia quis vel eros donec ac odio tempor. Urna nec tincidunt praesent semper feugiat nibh. Tortor aliquam nulla facilisi cras fermentum odio eu feugiat. Magna ac placerat vestibulum lectus mauris ultrices eros. Pulvinar proin gravida hendrerit lectus. Nisi scelerisque eu ultrices vitae auctor eu. Quis hendrerit dolor magna eget est lorem ipsum. At consectetur lorem donec massa sapien faucibus et molestie. Feugiat in fermentum posuere urna nec tincidunt praesent. Facilisis gravida neque convallis a cras. Massa id neque aliquam vestibulum morbi blandit. Faucibus purus in massa tempor nec feugiat nisl pretium fusce. Amet facilisis magna etiam tempor orci eu lobortis. Odio tempor orci dapibus ultrices in iaculis nunc sed augue. Nulla aliquet porttitor lacus luctus accumsan tortor posuere ac ut. Sit amet aliquam id diam maecenas. Rhoncus aenean vel elit scelerisque mauris pellentesque pulvinar pellentesque habitant. Urna nunc id cursus metus aliquam.

...

Iaculis nunc sed augue lacus viverra vitae congue eu consequat. Sodales ut eu sem integer vitae. Pulvinar mattis nunc sed blandit libero volutpat sed cras. Maecenas volutpat blandit aliquam etiam erat velit scelerisque in dictum. Ultricies integer quis auctor elit sed vulputate mi. Ornare suspendisse sed nisi lacus sed viverra tellus in. Quis commodo odio aenean sed adipiscing diam donec. Platea dictumst vestibulum rhoncus est pellentesque elit. Commodo nulla facilisi nullam vehicula ipsum a arcu cursus. Ut faucibus pulvinar elementum integer enim neque volutpat. Diam quis enim lobortis scelerisque fermentum dui.

>>>>

Overview of the global burden of mental health

Cite key literature – systematic reviews, Lancet series, Vikram’s work, DC3P findings, Walker et al. EMERALD health systems & mental health study, Mental Health Atlas from WHO, GBD for estimates of burden

Implications of this global burden for health systems

Highlight interconnectedness of mental health and physical health

Links of mental health to financial risk protection / vulnerabilities to poverty

Underscore under-investment in mental health systems, coupled with lack of data on mental health, which constrains research and policy

Also note that mental health is a truly global issue, impacting HICs and LMICs alike

Overview of Paper 1

Overview of Paper 2

Overview of Paper 3

### 2023-02-25 paper3.Rmd
Abstract and Introduction for Paper 3

paper3_abstract <- paste0("For searches related to depression, we observe a statistically significant slope change of -0.0304 percentage points per day (p < 0.0001, 99\\% CI: -0.0321– -0.0286) and a level shift of -0.184 (p < 0.0001, 99\\% CI: -0.226– -0.143), above and beyond changes in slope and level observed for searches for headaches. For anxiety, we observe a similar slope change decline of 0.0383 percentage points per day (p < 0.0001, 99\\% CI: -0.0400– -0.0366) and a level shift of -0.339 (p < 0.0001, 99\\% CI: -0.378– -0.299).  We do not observe significant variation of estimates across predominately Democrat or Republican voting states and counties, and our results are robust to alternative specifications of bandwidth, fixed effects, and data cleaning decisions.")

@@

## Introduction
It has been well-documented that throughout the pandemic, symptoms of depression and anxiety have markedly increased among US adults. While it is likely that the pandemic has a direct effect on worsened mental health status due to fear of contracting and spreading the virus, public health responses to the pandemic—including lockdowns, school closures, and social distancing measures—likely have an indirect effect as well. Existing research suggests that the effects of these measures on mental health vary, and may be protective, null, or even harmful. Studies of lockdowns in multiple countries, for example, point to worsened feelings of isolation and hopelessness exacerbated by isolation measures. By contrast, the distribution of safe and effective vaccines against COVID-19 has been shown to reduce symptoms of depression and anxiety by as much as XXX.

* Describe symptoms database and reference published literature using it to predict COVID-19 cases and deaths [@alruilyPredictionCovid19Transmission2022; @abbasAssociationsGoogleSearch2021; @saegnerForecastingSurveillanceCOVID192022]

A recently published evaluation of Google’s COVID-19 Search Trends Symptoms Dataset found strong correlation between search trends for depression and anxiety with indicators of mental health service utilization and need reported in the US Census Household Pulse Survey (HPS) with rates of emergency department visits for mental health condition as reported in the US Centers for Disease Control and Prevention’s (CDC) National Syndromic Surveillance Program (NSSP).[@vaidyanathanEvaluationInternetSearches2022]

controlled interrupted time series (CITS) 
interrupted time series (ITS)

* Motivation to explore vaccine announcement: evidence that vaccination itself reduced depression and anxiety -- do we see earlier reductions from an anticipatory effect?
* Motivation to analyze by polity: Vaccination became politically charged in the US, possible bias from impact of 2020 election on mental health searches
* Overview of approach: CITS using searches for headache as a control, adjusting for fixed effects, subgroup analysis by polity, analysis for 6 different mental health symptoms

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. A growing body of evidence has documented that political ideology can play a role in how receptive people are to vaccination; this has been highlighted throughout the COVID-19 pandemic, with people identifying as conservative being more likely to express skepticism of public health authorities and the scientific community and being more likely to express concerns about the safety, effectiveness, or necessity of vaccines. Our subgroup analysis comparing CITS effect estimates in blue vs. red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

@@

  combo_plot_1 <- cowplot::plot_grid(plot_1,
                                   plot_2,
                                   plot_3,
                                   # plot_4,
                                   # plot_5,
                                   # plot_6,
                                   # plot_7,
                                   # plot_8,
                                   # plot_9,
                                   align='v', ncol=1, axis='b') 
 print(combo_plot_1)

@@

@@

@@

While our ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. Our CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed our ITS analysis, with most conditions showing level and slope changes close to the null. For three conditions, our CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")`, all showing sustained declines. All three of these symptoms had relatively large slope changes in our ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "level")`, a decrease double in magnitude compared to the majority of estimates for other symptoms.

@@

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_depression, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for depression, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_depression}Density of Google Search trends for depression, by region and as a function of time. In each row, search density data for two symptoms are displayed: depression in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_mdd, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4, fig.height = 6.5, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for major depressive disorder (MDD), by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_mdd}Density of Google Search trends for MDD, by region and as a function of time. In each row, search density data for two symptoms are displayed: MDD in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r fig_international, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7.5, fig.height = 6.5, fig.align='center', out.width = '95%',  fig.scap="International CITS effect estimates and confidence intervals by mental health symptom.", out.extra='',  fig.cap="\\label{fig_international}International controlled interrupted time series effect estimates of changes in search density by mental health related symptom, reported across six nations (Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States). The top panel reports estimates of the level shift and the lower panel reports the estimates for the slope change. Effect estimates are represented as solid black lines, with the shaded regions above and below the lines showing the 95% confidence intervals. The solid black line across all panels reflects a null effect. MDD: Major depressive disorder."}

>>>>

@@

@@

  combo_plot_1 <- cowplot::plot_grid(plot_1,
                                   plot_2,
                                   plot_3,
                                   # plot_4,
                                   # plot_5,
                                   # plot_6,
                                   # plot_7,
                                   # plot_8,
                                   # plot_9,
                                   align='v', nrow=3, axis='b') 
 print(combo_plot_1)

@@

paper3_abstract <- paste0("For searches related to depression, we observe a statistically significant slope change of ",
                          function_dynamic_text_results("main","depression", "slope"), " and a level shift of ",
                          function_dynamic_text_results("main","depression", "intercept"),
                          ", above and beyond changes in slope and level observed for searches for headaches. For anxiety, we observe a similar slope change decline of",
                          function_dynamic_text_results("main","anxiety", "slope"), 
                          " and a level shift of ",
                          function_dynamic_text_results("main","anxiety", "intercept"),
 ". We do not observe significant variation of estimates across predominately Democrat or Republican voting states and counties, and our results are robust to alternative specifications of bandwidth, fixed effects, and data cleaning decisions.")

@@

## Introduction
The coronavirus disease 2019 (COVID-19) pandemic has had a profound effect on the mental health and wellbeing of individuals worldwide. In the United States, approximately four in ten adults reported experiencing symptoms of anxiety or depression in December 2020—a four-fold increase from January-June 2019.[@panchalImplicationsCOVID19Mental2021, p. 19] Studies from Australia,[@batterhamTrajectoriesDepressionAnxiety2021] China,[@ahmedEpidemicCOVID19China2020; @gaoMentalHealthProblems2020; @hwangLonelinessSocialIsolation; @leiComparisonPrevalenceAssociated2020; @liuPrevalencePredictorsPTSS2020; @qiuNationwideSurveyPsychological2020; @wangImmediatePsychologicalResponses2020; @wangStudyPublicPsychological2021; @wangPsychologicalDistressCoping2020; @zhangImpactCOVID19Pandemic2020] Denmark,[@sonderskovDepressiveStateDenmark] Europe,[@hajekPrevalenceDeterminantsProbable2022] Iran,[@moghanibashi-mansouriehAssessingAnxietyLevel2020] Ireland,[@hylandAnxietyDepressionRepublic2020]  Italy,[@mazzaNationwideSurveyPsychological2020; @mocciaAffectiveTemperamentAttachment2020] Nepal,[@samadarshiOnlineSurveyFactors2020] New Zealand,[@gasteigerDepressionAnxietyStress2021] Spain,[@gonzalez-sanguinoMentalHealthConsequences2020; @ozamiz-etxebarriaStressAnxietyDepression2020] Turkey[@ozdinLevelsPredictorsAnxiety2020],  the United States,[@jiaNationalStateTrends2021, p. 2020–2021; @kujawaExposureCOVID19Pandemic2020] and the United Kingdom[@dettmannInvestigatingPrevalenceAnxiety2022] report high levels of anxiety (6-51%), depression (15-48%), and post-traumatic stress disorder (7-54%) associated with aspects of the pandemic.  Globally, it is estimated that the prevalence of depression during the COVID-19 pandemic may be as high as 24%, with the prevalence of anxiety estimated at 21.3%.[@banko-ferranImpactCOVID19Mental2020; @castaldelli-maiaInvestigatingEffectNational2021] This striking burden has been described as a mental health crisis[@melimopoulosHowCOVID19Pandemic2021] and a pandemic in its own right.[@kwaiWhatPointYoung2021]

While COVID-19 continues to bring unprecedented challenges to individuals and societies around the world, the development and distribution of safe and effective vaccines against COVID-19 has fundamentally changed the trajectory of the pandemic. Beyond the physical benefits vaccines provide in reducing the risk of infection and associated morbidity and mortality, a growing body of evidence suggests that vaccination has also contributed to improved mental health.[@agrawalImpactCOVID19Vaccine2021; @chourpiliadisShorttermImprovementMental2023; @nguyenPsychologicalBenefitsCOVID192021; @rubinsteinEffectCOVID19Vaccinations2023] Research drawing from the United States Census Bureau’s Household Pulse Survey, for example, has found that the secondary benefits of the protective effects of vaccination may include up to a 30% reduction in symptoms of anxiety and depression in vaccinated adults.[@agrawalImpactCOVID19Vaccine2021]

While this literature suggests that that receiving a COVID-19 vaccine can improve mental health state, the potential mental health benefits of public health announcements regarding the development of vaccines have received comparatively little attention. In the context of a deadly pandemic, the announcement of a novel safe and effective vaccine could be a powerful signal, providing a psychological boost and hope that the worst might soon be mitigated. Studying how population mental health may have been impacted by the release of vaccine efficacy and safety trial data may provide insights into how emotional well-being changed throughout the COVID-19 pandemic in response to these signals, which, in turn, could inform public health communication and contribute to our understanding of how the public processes and responds to information about vaccines—a necessary understanding for increasing vaccine uptake and trust in public health responses.

A key obstacle to studying the effects of public health responses on population mental health in the US, however, is a lack of daily surveillance data on mental health indicators; a potential marker of population mental health is the use of Internet search data, which offers the potential for real-time, large-scale analysis and identification of evolving mental health issues. A recently published evaluation of Google’s COVID-19 Search Trends Symptoms Dataset found strong correlation between search trends for depression and anxiety with indicators of mental health service utilization and need reported in the US Census Household Pulse Survey (HPS) with rates of emergency department visits for mental health condition as reported in the US Centers for Disease Control and Prevention’s (CDC) National Syndromic Surveillance Program (NSSP).[@vaidyanathanEvaluationInternetSearches2022] If taken as a reasonable proxy for mental health complaints, search data can provide valuable information on how population mental health reacts to real-time events and evolves over different shocks, enabling policymakers to better respond to mental health needs and assess reactions to public health efforts.

Our study examines how news regarding the imminent availability of multiple safe and effective COVID-19 vaccines in November 2020 may have impacted population mental health even before these vaccines were made available to the public. Using Google search density data, we employed an interrupted time series approach to investigate how these announcements may have affected search trends for mental health symptoms in the United States and in five other countries. Owing to emerging evidence that political ideology may be a predictor of confidence in public health and vaccine hesitancy[@debusPoliticalIdeologyVaccination2021; @baumgaertnerInfluencePoliticalIdeology2018; @stolerSociopoliticalPsychologicalCorrelates2022] (with people identifying as conservative being more likely to express skepticism of public health authorities and the scientific community and being more likely to express concerns about the safety, effectiveness, or necessity of vaccines), we further explored whether changes in search trends for mental health symptoms varied across states and counties that predominantly voted Democrat versus Republic in the 2020 president election. Our findings suggest both immediate and sustained declines in search density for searches related to anxiety, depression, and major depressive disorder; these declines were observed both in the U.S. and abroad, and were found to be consistent irrespective of political ideology.

The second subgroup analysis applied the CITS approach to the 20% most Republican-voting localities (hereafter referred to as red states/counties) and to the 20% most Democratic-voting localities (hereafter referred to as blue states/counties) based on 2020 presidential election vote share. Our subgroup analysis comparing CITS effect estimates in blue vs. red states aimed to investigate whether the initial response to imminent safe and effective COVID-19 vaccines showed evidence of heterogeneity by predominant political ideology, under the hypothesis that if conservative voters were more likely to be averse to vaccination, any impact on mental health related symptom search density from the announcement of upcoming vaccines would be lesser in red states and counties compared to blue ones.

@@

While our ITS analysis provided suggestive evidence that search density for mental health symptoms might have changed in the post-announcement period, it also highlighted that searches may have generally changed across other diseases at the same time. Our CITS estimates of each mental health symptom (Figure \ref{figure_cits_outputs}) using searches for headaches as a comparison broadly echoed our ITS analysis, with most conditions showing level and slope changes close to the null. For three conditions, our CITS analysis showed relatively large slope changes with searches for anxiety `r function_dynamic_text_results("main","anxiety", "slope", value_include = "allnumber")`, depression `r function_dynamic_text_results("main","depression", "slope", value_include = "allnumber")`, and MDD `r function_dynamic_text_results("main","major_depressive_disorder", "slope", value_include = "allnumber")`, all showing sustained declines. All three of these symptoms had relatively large slope changes in our ITS analysis. In addition, the CITS level shift (i.e., the relative immediate effect) for anxiety was `r function_dynamic_text_results("main","anxiety", "intercept")`, a decrease double in magnitude compared to the majority of estimates for other symptoms.

@@

```{r figures_cits_data_anxiety, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4.5, fig.height = 7, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for anxiety, by region, as a function of time.", out.extra='', fig.cap="\\label{figures_cits_data_anxiety}Density of Google Search trends for anxiety, by region and as a function of time. In each row, search density data for two symptoms are displayed: anxiety in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_depression, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4.5, fig.height = 7, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for depression, by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_depression}Density of Google Search trends for depression, by region and as a function of time. In each row, search density data for two symptoms are displayed: depression in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r figures_cits_data_mdd, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 4.5, fig.height = 7, fig.align='center', out.width = '99%', fig.scap="Density of Google Search trends for major depressive disorder (MDD), by region, as a function of time.", out.extra='',  fig.cap="\\label{figures_cits_data_mdd}Density of Google Search trends for MDD, by region and as a function of time. In each row, search density data for two symptoms are displayed: MDD in blue and headaches in grey. In the first column, data from US states are shown, where each point represents the search density of a specific symptom on a specific date in a specific state. In the second column, data from US counties are shown; because of the large number of counties reporting data, county-level search data are shown as daily boxplots. In the third column, national data from Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States are shown. Each graph shows the density of symptom searches on the y-axis, while time is centered and displayed as days prior to and following November 16, 2020 on the x-axis. The solid lines show predicted lines of best fit under a basic controlled interrupted time series (CITS) model (i.e., no fixed effects). US: United States."}

@@

```{r fig_international, include = TRUE, echo=FALSE, warnings = FALSE, fig.width = 7.5, fig.height = 6.5, fig.align='center', out.width = '95%',  fig.scap="International CITS effect estimates and confidence intervals by mental health symptom.", out.extra='',  fig.cap="\\label{fig_international}International controlled interrupted time series effect estimates of changes in search density by mental health related symptom, reported across six nations (Australia, the United Kingdom, Ireland, New Zealand, Singapore, and the United States). The top panel reports estimates of the level shift and the lower panel reports the estimates for the slope change. Effect estimates are represented as solid black lines, with the shaded regions above and below the lines showing the 95\\% confidence intervals. The solid black line across all panels reflects a null effect. MDD: Major depressive disorder."}

### 2023-02-25 thesis.Rmd
Abstract and Introduction for Paper 3

# Maps in Chapter 1 need to be improved to eliminate errors in cartographic resolution
# Separate bibliography for each section
# Numbered figures and tables by section
# Check that sequential articles are represented by a dash
# Table of figures and tables
# Acronyms with acronym down
# Institutional authors need to be fixed in bibliography

>>>>

# Acronyms with acronym down
# Separate bibliography for each section
# Numbered figures and tables by section

# Check that sequential articles are represented by a dash
# Table of figures and tables
# Institutional authors need to be fixed in bibliography
